Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-07 paper/ms.tex
update ms

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 1-4\,Gyr, and provides more complex asymmetric posteriors at earlier
times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
clock has a rich history
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  In the
context of Kepler, empirical models have been previously applied to
derive rotation-based ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  There have been
major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The extent of the Kepler Object of Interest catalog, and our vetting of false
    positives within it;
  %
  \item Available measurements of stellar rotation periods
    in open clusters
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation through time
    \citep{Bouma_2023}.  For Sun-like stars, the new model is broadly
    consistent with older calibrations \citep[e.g.][]{Mamajek_2008}.
    For late G dwarfs and K dwarfs, the age calibration
    differs by factors of a few;
  %
  \item Our ability to identify binary stars, which can be problematic
    for estimates of rotation-based ages.
\end{enumerate}
Today, FGK star rotation-based ages are
empirically tied to the cluster age scale, and
are limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.

@@

\subsubsection{New Young Systems}
%
Whenever both rotation and lithium find a star to be in a
narrowly-constrained age range, confidence in the true age
increases, because the two techniques are independent.
The region of greatest overlap between rotation and lithium ages in
the spaces of $P_{\rm rot}$, $EW_{\rm Li}$, and $T_{\rm eff}$ is
dictated by the timescale for lithium depletion: for stars with
temperatures between 4500-5500\,K with ages $\lesssim$500\,Myr, we can
hope to obtain independent two-sided constraints
\citep{Bouma_2023,Jeffries_2023}.

The following systems, to our knowledge, are being
well-constrained for the first time in this work.  Here we limit
discussion to systems with $t_{\rm gyro}$$<$500\,Myr not in the clusters
mentioned above.
{\bf FIXME TODO: this section is boring.  come up with a better way to do this,
and maybe automate}

@@

{\it Age Accuracy}: The fact that only $\approx$90\% of the
rotation-based and lithium-based ages agree suggests that in cases of
disagreement, either one or both could be inaccurate.

>>>>

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 1-4\,Gyr, and provides more complex asymmetric posteriors at earlier
times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
clock has a rich history
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  In the
context of Kepler, empirical models have been previously applied to
derive rotation-based ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  ``Gyro-kinematic'' ages
have also been reported \citep{2021AJ....161..189L}.
There have been
major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The extent of the Kepler Object of Interest catalog, and our vetting of false
    positives within it;
  %
  \item Available measurements of stellar rotation periods
    in open clusters
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation through time
    \citep{Bouma_2023}.  For Sun-like stars, the new model is broadly
    consistent with older calibrations \citep[e.g.][]{Mamajek_2008}.
    For late G dwarfs and K dwarfs, the age calibration
    differs by factors of a few;
  %
  \item Our ability to identify binary stars, which can be problematic
    for estimates of rotation-based ages.
\end{enumerate}
Today, FGK star rotation-based ages are
empirically tied to the cluster age scale, and
are limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.

@@

\subsubsection{New Young Systems}
%
Whenever both rotation and lithium find a star to be in a
narrowly-constrained age range, confidence in the true age
increases, because the two techniques are independent.
The region of greatest overlap between rotation and lithium ages in
the spaces of $P_{\rm rot}$, EW$_{\rm Li}$, and $T_{\rm eff}$ is
dictated by the timescale for lithium depletion: for stars with
temperatures between 4500-5500\,K with ages $\lesssim$500\,Myr, we can
hope to obtain independent two-sided constraints
\citep{Bouma_2023,Jeffries_2023}.

The following systems, to our knowledge, are being
well-constrained for the first time in this work.  Here we limit
discussion to systems with $t_{\rm gyro}$$<$500\,Myr not in the clusters
mentioned above.
{\bf FIXME TODO: next two paragraphs are boring.  come up with a better way to do this,
and maybe automate}

@@

{\it Age Accuracy and Quality Flagging}: The fact that only $\approx$90\% of the
rotation-based and lithium-based ages agree suggests that in cases of
disagreement, either one or both could be inaccurate.
